---
title: "EDS241: Assignment 4"
author: "Clarissa Boyajian"
date: '`r format(Sys.time(), "%m/%d/%Y")`'
output: 
  pdf_document:
    toc: false
    number_sections: yes
header-includes:
  - \setlength{\parindent}{1em}
  - \usepackage{float}
--- 
  
``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}
# set default chunk options
knitr::opts_chunk$set(fig.width = 4, fig.height = 3, 
                      echo = TRUE, message = FALSE, warning = FALSE)

# load packages
packages = c("stargazer", "here", "tidyverse", "janitor", 
           "cowplot", "tinytex", "datasets", "estimatr", "tibble", "car", 
           "readxl", "huxtable", "AER", "knitr")

for (i in packages) {
  if (require(i, character.only = TRUE) == FALSE) {
    install.packages(i,repos = 'http://cran.us.r-project.org')
  }
  else {
    require(i, character.only = TRUE)
  }
}

options(scipen = 999) # not scientific notation
```

# Assignment 4

\noindent In this assignment we are estimating the price elasticity of demand for fresh sardines across 56 ports located in 4 European countries with monthly data from 2013 to 2019. 

## Load and clean data

\noindent The following code loads and cleans the data and adds a column of `log()` values of the price and volume sold of sardines.

```{r}
data_sardines <- read.csv("data/EU_sardines.csv") %>% 
  clean_names() %>% 
  mutate(log_price_euro_kg = log(price_euro_kg),
         log_volume_sold_kg = log(volume_sold_kg))
```


## Question A
\noindent Estimate a bivariate regression of `log(volume_sold_kg)` on `log(price euro_kg)`. What is the price elasticity of demand for sardines? Test the null hypothesis that the price elasticity is equal to -1.

\noindent The code chunk below runs a bivariate regression of `log(volume_sold_kg)` on `log(price euro_kg)`:

```{r}
model_1 <- lm_robust(data = data_sardines,
                     log_volume_sold_kg ~ log_price_euro_kg)


model_1_table <- tidy(model_1) %>% 
  select(term, estimate, std.error, p.value, conf.low, conf.high) %>% 
  kable()

model_1_table
```

\noindent \textbf{a.} The estimated price elasticity of demand for sardines regressing log volume on log price is `r round(model_1[[1]][[2]], 2)`. We can say with 95% confidence that the intervals of `r round(model_1$conf.low[[2]], 2)` to `r round(model_1$conf.high[[2]], 2)` contains the true $\beta_1$ (aka price elasticity). Because this range does not include -1, we can reject the null hypothesis that the price elasticity is equal to -1.


## Question B
\noindent Like in Lecture 8 (see the IV.R script), we will use wind_m_s as an instrument for `log(price_euro_kg)`. To begin, estimate the first-stage regression relating `log(price_euro_kg)` to `wind_m_s.` Interpret the estimated coefficient on wind speed. Does it have the expected sign? Also test for the relevance of the instrument and whether it is a “weak” instrument by reporting the proper F-statistic. 

\noindent The code chunk below estimates the first-stage regression relating `log_price_euro_kg` to `wind_m_s`:

```{r}
# first stage regression
model_2 <- lm_robust(data = data_sardines,
                     log_price_euro_kg ~ wind_m_s)

huxreg("log_price_euro_kg" = model_2)
```

\noindent The code chunk below calculates the proper F-statistic:

```{r}
model_2_fstat <- linearHypothesis(model_2, c("wind_m_s = 0"), 
                                  white.adjust = "hc2")

huxtable(model_2_fstat)
```

\noindent \textbf{b.} The estimated coefficient on wind speed is `r round(model_2[[1]][2], 4)`, meaning that for 1 meter per second increase in wind speed there is an increase of `r round(model_2[[1]][2], 4)` in the log price of sardines. The sign is positive, which we would expect because higher wind speeds may lead to a lower total supply of sardines for a given day. Because wind speed doesn't effect the demand for sardines, the log price goes up when supply decreases but the demand stays the same. The F-statistic is equal to `r round(model_2_fstat[[3]][2], 2)`, which is greater than 10. This means that our instrument (wind speed) is not weak.

\newpage

## Question C
\noindent Estimate the TSLS estimator of the price elasticity of demand for sardines using `wind_m_s` as an instrument for `log(price_euro_kg)`. What is the estimated price elasticity of demand for sardines?

\noindent The code chunk below estimates the TSLS estimator of the price elasticity of demand for sardines using `wind_m_s` as an instrument for `log_price_euro_kg`:

```{r}
TSLS_1 <- ivreg(data = data_sardines,
                log_volume_sold_kg ~ log_price_euro_kg | wind_m_s)

huxreg("log_volume_sold_kg" = TSLS_1)
```

\noindent \textbf{c.} The estimated price elasticity of demand for sardines using wind as an instrument of log price is `r round(TSLS_1[[1]][[2]], 2)`.

## Question D
\noindent Repeat the exercise in (c), but include fixed effects for each year, month, and country. Report the estimated price elasticity of demand and the F-statistic testing for relevant and non-weak instruments.

\noindent The code chunk below The code chunk below estimates the TSLS estimator of the price elasticity of demand for sardines using `wind_m_s` as an instrument for `log_price_euro_kg` including a fixed effects for each year, month, and country:

```{r}
TSLS_2 <- ivreg(data = data_sardines,
                log_volume_sold_kg ~ log_price_euro_kg + as.factor(country) + 
                  as.factor(year) + as.factor(month) | 
                  as.factor(country) + as.factor(year) + 
                  as.factor(month) + wind_m_s)

TSLS_2_huxreg <- huxreg(TSLS_2)
```

\newpage 

```{r}
restack_across(TSLS_2_huxreg, 21)
```

\newpage

\noindent The code chunk below calculates the proper F-statistic:

```{r}
# calculate F-statistic
model_3 <- lm_robust(data = data_sardines,
                     log_price_euro_kg ~ wind_m_s + as.factor(country) + 
                       as.factor(year) + as.factor(month))

model_3_fstat <- linearHypothesis(model_3, c("wind_m_s = 0"), 
                                  white.adjust = "hc2")
huxtable(model_3_fstat)
```

\noindent \textbf{d.} The estimated price elasticity of demand for sardines using wind as an instrument of log_price and including fixed effects for each year, month, and country is `r round(TSLS_2[[1]][[2]], 2)` and the F-statistic for the instrument of windspeed is `r round(model_3_fstat[[3]][2], 2)`. 




